Rapid-response Bayesian sky localization for electromagnetic follow-up of gravitational-wave candidates

### 2013-03-07 ms.tex
intro text and some references

Ground-based gravitational-wave interferometers are entering the advanced detector era, positioning themselves to make profound discoveries about the universe. The advanced LIGO and Vigo detectors will be part of a worldwide network that also includes KAGRA and hopefully LIGO-India. Coalescing neutron star binaries are among the most likely sources, with event rates of $~ 50/\mathrm{yr}$ (check that number!). In addition to being an efficient source of gravitational waves, the neutron star may be tidally shredded before merging, providing fuel for an electromagnetic counterpart. The short-lived accretion flow may power a collimated relativistic jet, resulting in a short gamma-ray burst and an X-ray afterglow if it is aligned with the line of sight. An optical afterglow may follow minutes to hours later as the jet plows into the interstellar medium. As surrounding neutron-rich ejecta decay radioactively, an optical `kilonova' may be visible after $~$1 day. Finally, as the ejecta plow into the interstellar medium, a faint radio afterglow will occur on a timescale of years.

There is therefore a strong case for performing electromagnetic followups of these sources of gravitational waves. Indeed, the final science run of the intitial LIGO and Virgo instruments saw the first joint search for gravitational waves from compact binaries and their electromagnetic counterparts....

In this paper we present a method of rapid sky localization that makes use of Bayesian methods. It differs from...

We have developed a rapid-response sky localization algorithm called \ac{BAYESTAR}%
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, gravitational-wave detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `T' in the acronym, because then it would be called BAYES\nobreakdashes-TART, which would sound ridiculous.}%

>>>>

Ground-based gravitational-wave interferometers are entering the advanced detector era, positioning themselves to make profound discoveries about the universe. The advanced LIGO and Vigo detectors will be part of a worldwide network that also includes KAGRA and hopefully LIGO-India. Coalescing neutron star binaries are among the most likely sources, with event rates of $~ 50/\mathrm{yr}$ (check that number!)\cite{rates}. In addition to being an efficient source of gravitational waves, the neutron star may be tidally shredded before merging, providing fuel for an electromagnetic counterpart. The short-lived accretion flow may power a collimated relativistic jet, resulting in a short gamma-ray burst and an X-ray afterglow if it is aligned with the line of sight. An optical afterglow may follow minutes to hours later as the jet plows into the interstellar medium. As surrounding neutron-rich ejecta decay radioactively, an optical `kilonova' may be visible after $~$1 day \cite{metzger:2010}. Finally, as the ejecta plow into the interstellar medium, a faint radio afterglow could occur on a timescale of weeks to years \cite{Nakar:2011cw}.

There is therefore a strong case for performing electromagnetic followups of these sources of gravitational waves. Indeed, the final science run of the intitial LIGO and Virgo instruments saw the first joint search for gravitational waves from compact binaries and their electromagnetic counterparts \cite{abadie2012first}.  As we prepare for the next generation of these searches there is a need for determining the location of the source as precisely as possible in as little time as possible.

In this paper we present a rapid and accurate method of sky localization that makes use of Bayesian methods. It differs from existing techniques is several important ways.  In the first place, we fix the intrinsic parameters to their maximum likelihood estiamtes, as provided by the detection pipeline.  This reduces the dimensionality of the parameter we need to search.  In addition, the technique takes as its input the maximum likelihood estiamtes of the extrinsic parameters, instead of the $h(t)$ time series.  Finally, instead of of using Markov Chain Monte Carlo or some similar method for statistical sampling, we make use of a deterministic quadrature scheme.   We call this technique \ac{BAYESTAR}%
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, gravitational-wave detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `T' in the acronym, because then it would be called BAYES\nobreakdashes-TART, which would sound ridiculous.}%
.
